---
title: "NHANES case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHANES case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

library(table.glue)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(tibble)
library(gt, warn.conflicts = FALSE)

```

# Overview

This vignette shows how to go from data to table using `table.glue` and friends. For this vignette, we will analyze data from The US National Health and Nutrition Examination Survey (NHANES), 2013-2018. 

NHANES was designed to assess the health and nutritional status of the non-institutionalized US population and was conducted by the National Center for Health Statistics of the Centers for Disease Control and Prevention. Since 1999-2000, NHANES has been conducted in two-year cycles using a multistage probability sampling design to select participants. Each cycle is independent with different participants recruited.

# NHANES Blood pressure measurements

The same protocol was followed to measure systolic and diastolic blood pressure (SBP and DBP) in each NHANES cycle. After survey participants had rested 5 minutes, their BP was measured by a trained physician using a mercury sphygmomanometer and an appropriately sized cuff. Three BP measurements were obtained at 30 second intervals.

```{r}

# see ?table.glue::nhanes for description of all columns

glimpse(nhanes)

```

# Making the analysis data

We will make a table showing SBP and DBP for subgroups based on sex and race over the three NHANES cycles in `nhanes`. But first need to filter to include rows where the survey participant

- completed both the NHANES interview and exam

- is 20 years or older

- is not pregnant

- is currently using medication to lower their BP

- does not have missing data for SBP or DBP

```{r}

nhanes_analysis <- nhanes %>% 
  filter(exam_status == 'interview and exam', 
         age >= 20, 
         pregnant == 'no' | is.na(pregnant),
         meds_bp == 'yes') %>% 
  select(exam, sex, race_ethnicity, bp_sys_mmhg, bp_dia_mmhg) %>% 
  na.omit()

nhanes_analysis

```

# Making the summary data

For each NHANES exam and sex/race/ethnicity group, we would like to compute the mean and standard deviation of BP. We'll use `dplyr` for this.

```{r}

nhanes_bp_summary <- nhanes_analysis %>% 
  group_by(exam, sex, race_ethnicity) %>% 
  summarize(
    across(
      .cols = c(bp_sys_mmhg, bp_dia_mmhg),
      .fns = list(
        mean = ~ mean(.x),
        sd = ~ sd(.x)
      )
    )
  )

nhanes_bp_summary

```

# Making the summary strings

With our summary data, we can now create the data that will be passed to our table. This is where `table.glue` comes into the picture.

```{r}

rspec <- round_spec() %>% 
  round_half_even() %>% 
  round_using_magnitude(
    digits = c(2, 1,  1,   0),
    breaks = c(1, 10, 100, Inf)
  )

nhanes_bp_strings <- ungroup(nhanes_bp_summary) %>% 
  transmute(
    exam,
    sex, 
    race_ethnicity,
    sbp = table_glue("{bp_sys_mmhg_mean} ({bp_sys_mmhg_sd})", rspec=rspec),
    dbp = table_glue("{bp_dia_mmhg_mean} ({bp_dia_mmhg_sd})", rspec=rspec)
  )

```

# Making the table object

Making a table out of `nhanes_tbl_strings` is fairly straightforward, but does require a fair bit of code to recode variables and label things. 

```{r}

nhanes_bp_table <- nhanes_bp_strings %>% 
  mutate(
    race_ethnicity = recode(
      race_ethnicity,
      'hispanic' = "Mexican American",
      'asian' = "Non-Hispanic Asian",
      'black' = "Non-Hispanic Black",
      'white' = "Non-Hispanic White",
      'other' = "Other Race - Including Multi-Racial"
    ),
    sex = recode(
      sex, 
      'female' = 'Female survey participants', 
      'male' = 'Male survey participants'
    ),
    exam = recode(
      exam,
      'exam_2013_2014' = '2013-2014',
      'exam_2015_2016' = '2015-2016',
      'exam_2017_2018' = '2017-2018'
    )
  ) %>%
  pivot_wider(names_from = 'exam',
              values_from = c('sbp', 'dbp')) %>% 
  gt(rowname_col = 'race_ethnicity', 
     groupname_col = 'sex') %>% 
  cols_label(
    "sbp_2013-2014" = "2013-2014",
    "sbp_2015-2016" = "2015-2016",
    "sbp_2017-2018" = "2017-2018",
    "dbp_2013-2014" = "2013-2014",
    "dbp_2015-2016" = "2015-2016",
    "dbp_2017-2018" = "2017-2018"
  ) %>% 
  cols_align('center') %>% 
  tab_stubhead(label = 'Race / ethnicity') %>% 
  tab_spanner(label = 'Systolic BP, mm Hg',
              columns = c("sbp_2013-2014",
                          "sbp_2015-2016",
                          "sbp_2017-2018")) %>% 
  tab_spanner(label = 'Diastolic BP, mm Hg',
              columns = c("dbp_2013-2014",
                          "dbp_2015-2016",
                          "dbp_2017-2018")) %>% 
  tab_header(
    title = "Systolic and diastolic blood pressure for female and male participants in the US National Health and Nutrition Examination Survey from 2013 - 2018", 
    subtitle = "Table values are mean (standard deviation)"
  )

nhanes_bp_table

```

# Making the inline object

Now we just need to summarize the table, which can be done using an inline table object. Just like the table, our inline object is created using `nhanes_bp_strings`

```{r}

nhanes_bp_inline <- nhanes_bp_strings %>% 
  as_inline(tbl_variables = c('exam', 'sex', 'race_ethnicity'),
            tbl_value = c('sbp', 'dbp'))

# Now you can freely access any cell in the table
nhanes_bp_inline$exam_2013_2014$male$black$sbp
nhanes_bp_inline$exam_2017_2018$male$black$sbp 

```

# Using the inline object

This part is easy: "Among survey participants who were black and female, the mean (standard deviation) of systolic blood pressure increased from `r nhanes_bp_inline$exam_2013_2014$female$black$sbp` in 2013-2014 to `r nhanes_bp_inline$exam_2017_2018$female$black$sbp` in 2017-2018."

But what if you don't want to report the standard deviations? These are helpful in the table but less so in the text. To get rid of these, you can use the bracket helper functions, e.g., `bracket_drop()` and `bracket_extract()`. 


```{r}

# Drop the bracket
bracket_drop(nhanes_bp_inline$exam_2013_2014$male$black$sbp)
# Get the bracket
bracket_extract(nhanes_bp_inline$exam_2017_2018$male$black$sbp)

```

Let's re-write our sentence more directly: "Among survey participants who were black and female, the mean systolic blood pressure increased from `r bracket_drop(nhanes_bp_inline$exam_2013_2014$female$black$sbp)` mm Hg in 2013-2014 to `r bracket_drop(nhanes_bp_inline$exam_2017_2018$female$black$sbp)` mm Hg in 2017-2018."


# Summary

This case study has covered three big ideas: 

1. Making summary data

1. Making summary strings 

1. Making the table and inline object

The workflow covered here is designed to fit almost any table and work with any package that creates tables. Happy tabulating!
