---
title: "Setting a default rounding specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting a default rounding specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

If you ever get tired of having to write `table_glue(..., rspec = your_rspec)` and would rather just write  `table_glue(...)`, you should find this vignette helpful.

## Options hierarchy

When you call a function like `table_glue()` or `table_value()`, you may supply your own rounding specification or use a default rounding specification. The default rounding specification is based on the global options of your current R session. So to make a particular rounding specification the default specification, all you need to do is set your global options using the specification of your choice. 

```{r setup}
library(table.glue)
```

## Step 1 - Make your rounding specification

Any rounding specification will do.

```{r}

rspec <- round_spec() 
rspec <- round_using_decimal(rspec, digits = 5)

```

## Step 2 - Modify specification names

Modify each name so that it starts with 'table.glue'. This protects you from mistakenly changing other package options.

```{r}

names(rspec) <- paste('table.glue', names(rspec), sep = '.')

```

## Step 3 - Pass to options.

All you need to do now is use the `options()` function. 

```{r}

options(rspec)

table_value(pi)

```

## Step 4 - Revert!

Did you make a mistake? We all do. You can revert the options back to normal by setting `force_default = TRUE` in a new `round_spec()` function, and then repeating steps 2 and 3.

```{r}

back_to_normal <- round_spec(force_default = TRUE)

names(back_to_normal) <- paste('table.glue', names(back_to_normal), sep = '.')

options(back_to_normal)

table_value(pi)

```

